---
output: html_document
editor_options: 
  chunk_output_type: console
---
# The Targeted Learning Roadmap {#intro}

A central goal of the Targeted Learning statistical paradigm is to estimate
scientifically relevant parameters in realistic (usually nonparametric) models.

## The Statistical Model
Assume we have an i.i.d. sample of confounders, a binary intervention of interest, and an outcome, or are observed data is  $O = (W, A, Y)$. The distribution of the observed data may
be factorized as follows: $P(O) = P(W, A, Y) = P(W)P (A \mid W) P(Y \mid A, W)$. To
estimate a parameter of interest, a researcher need not necessarily be able to
specify these whole or conditional distributions. Rather, each estimator only
requires that certain parts of the distribution be known; for example, some may
require estimates of $\mathbb{E}(Y \mid A, W)$, the mean of $Y$ within subgroups
$(A, W)$, or the regression of the outcome on the exposure and confounders.

At this stage in the roadmap, the researcher must specify a choice of
statistical model to be used in estimating $\mathbb{E}(Y \mid A, W)$ or other
elements of the probability distribution needed to estimate the parameter of
interest. Here, _statistical model_ means any constraints on the model form that
may be imposed by knowledge about the data-generating process -- that is, known
aspects of how the data were generated. Typically, the _true model_ is a very
large model, placing few constraints, if any, on the data-generating
distribution, or a semi-parametric model. With few constraints on the
data-generating distribution, and a potentially large number of covariates,
data-adaptive, machine-learning approaches remain the only practical option for
estimating components of the likelihood. The remainder of this course concerns
how to do this as efficiently and robustly as possible, depending on the goal of
the analysis.

## The Causal Model

The next step in the roadmap is to use a causal framework to formalize the
experiment and thereby define the parameter of interest. Causal graphs are one
useful tool to express what we know about the causal relations among variables
that are relevant to the question under study [@pearl2009causality]. 

Ignoring error terms, we will assume the following ordering of the variables in $O$.

```{r, echo=F, eval=T}
library(visNetwork)
nodes <- data.frame(id=c("W","A","Y"))
nodes$label <- nodes$id
edges <- data.frame(from=c("W","W","A"),to=c("A","Y","Y")) 
network <- visNetwork(nodes,edges, height="300px",width="200px") %>% 
  visEdges(arrows=list(to=TRUE))  %>% 
  visLayout(randomSeed=25)
print(network)
```

While
directed acyclic graphs (DAGs - like above) provide a convenient means by which to visualize
causal relations between variables, the  causal relations among variables
can be represented via a set of structural equations:
\begin{align*}
  W &= f_W(U_W) \\
  A &= f_A(W, U_A) \\
  Y &= f_Y(W, A, U_Y),
\end{align*}
where $U_W$, $U_A$, and $U_Y$ represent the unmeasured exogenous background
characteristics that influence the value of each variable. In the NPSEM, $f_W$,
$f_A$ and $f_Y$ denote that each variable (for $W$, $A$ and $Y$, respectively)
is a function of its parents and unmeasured background characteristics, but one typically has little information about  particular functional constraints (e.g., linear, logit-linear, only one interaction, etc.). For this
reason, they are called non-parametric structural equation models (NPSEMs). The
DAG and set of nonparametric structural equations represent exactly the same
information and so may be used interchangeably.

## The Parameter of Interest

The first hypothetical experiment we will consider is assigning exposure to the
whole population and observing the outcome, and then assigning no exposure to
the whole population and observing the outcome. On the nonparametric structural
equations, this corresponds to a comparison of the outcome distribution in the
population under two interventions:

1. $A$ is set to $1$ for all individuals, and
2. $A$ is set to $0$ for all individuals.

These interventions imply two new nonparametric structural equation models. For
the case $A = 1$, we have
\begin{align*}
  W &= f_W(U_W) \\
  A &= 1 \\
  Y(1) &= f_Y(W, 1, U_Y),
\end{align*}
and for the case $A=0$,
\begin{align*}
  W &= f_W(U_W) \\
  A &= 0 \\
  Y(1) &= f_Y(W, 0, U_Y).
\end{align*}

In these equations, $A$ is no longer a function of $W$ because we have
intervened on the system, setting $A$ deterministically to either of the values
$1$ or $0$. The new symbols $Y(1)$ and $Y(0)$ indicate the outcome variable in
our population if it were generated by the respective NPSEMs above; these are
often called _counterfactuals_. The difference between the means of the outcome
under these two interventions defines a parameter that is often called the
"average treatment effect" (ATE), denoted
\begin{equation}\label{eqn:ate}
  ATE = \mathbb{E}_X(Y(1)-Y(0)),
\end{equation}
where $\mathbb{E}_X$ is the mean under the theoretical (unobserved) full data
$X = (W, Y(1), Y(0))$.

## Identifiability

Because we can never observe both $Y(0)$ (the counterfactual outcome when $A=0$)
and $Y(1)$, we cannot estimate \ref{eqn:ate} directly. Instead, we have to make
assumptions under which this quantity may be estimated from the observed data
$O \sim P_0$ under the data-generating distribution $P_0$. Fortunately, given
the causal model specified in the NPSEM above, we can, with a handful of
untestable assumptions, estimate the ATE, even from observational data. These
assumptions may be summarized as follows

1. The causal graph implies $Y(a) \perp A$ for all $a \in \mathcal{A}$, which
   is the _randomization_ assumption. In the case of observational data, the
   analogous assumption is _strong ignorability_ or _no unmeasured confounding_
   $Y(a) \perp A \mid W$ for all $a \in \mathcal{A}$;
2. Although not represented in the causal graph, also required is the assumption
   of no interference between units, that is, the outcome for unit $i$ $Y_i$ is
   not affected by exposure for unit $j$ $A_j$ unless $i=j$;
3. _Consistency_ of the treatment mechanism is also required, i.e., the outcome
   for unit $i$ is $Y_i(a)$ whenever $A_i = a$, an assumption also known as "no
   other versions of treatment";
4. It is also necessary that all observed units, across strata defined by $W$,
   have a bounded (non-deterministic) probability of receiving treatment --
   that is, $0 < P_0(A = a \mid W) < 1$ for all $a$ and $W$). This assumption is
   referred to as _positivity_.

_Remark_: Together, (2) and (3), the assumptions of no interference and
consistency, respectively, are jointly referred to as the *stable unit
treatment value assumption* (SUTVA).

Given these assumptions, the ATE may be re-written as a function of $P_0$,
specifically
\begin{equation}\label{eqn:estimand}
  ATE = \mathbb{E}_0(Y(1) - Y(0)) = \mathbb{E}_0
    \left(\mathbb{E}_0[Y \mid A = 1, W] - \mathbb{E}_0[Y \mid A = 0, W]\right),
\end{equation}
or the difference in the predicted outcome values for each subject, under the
contrast of treatment conditions ($A = 0$ vs. $A = 1$), in the population,
averaged over all observations. Thus, a parameter of a theoretical "full" data
distribution can be represented as an estimand of the observed data
distribution. Significantly, there is nothing about the representation in
\ref{eqn:estimand} that requires parameteric assumptions; thus, the regressions
on the right hand side may be estimated freely with machine learning. With
different parameters, there will be potentially different identifiability
assumptions and the resulting estimands can be functions of different components
of $P_0$. We discuss several more complex estimands in later sections of this
workshop.

## Estimators: SuperLearning and Targeted Maximum Likelihood


Although we will discuss more in later sections, the goals of the estimators we
desire should be that, among sensible (asymptotically consistent, regular)
estimators,

1. the estimator be asymptotically efficient in the statistical model of
   interest, and
2. the estimator can be constructed for finite-sample performance improvements,
   relative to other estimators in the same class.

These priniciples guide our approach to estimation: Superlearning for prediction (more generally density estimation) and TMLE for estimation of our intervention parameters of interest. 

### SuperLearning
- There is no universally optimal machine learning algorithm (exceptions exist - HAL).
- This is true empirically, when we have test different algorithms on actual data and looked at the performance (e.g., MSE of prediction)
- For some data, one needs learners that can model a complex function.
- For others, typically caused by noise or insufficient sample size, a simple, parametric model might fit the best.
- SuperLearner, an ensemble learner, solves this issue, by allowing a convex combination of learners from the simplest (intercept-only) to most complex (neural nets, random forests, SVM, etc.).
- It works by using cross-validation in a manner which guarantees that the resulting fit will be as good as possible, given the learners provided (note, even a convex combination of poor learners can sometimes result in good fit, though better to have good candidates).

\includegraphics[width=5in]{vs.pdf}]




### TMLE
<object data="NatureSlides.pdf" width="750px" height="750px">
    <embed src="NatureSlides.pdf">
    </embed>
</object>


### Inference

The estimators we discuss are asymptotically linear, meaning that the difference
in the estimate $\Psi(P_n)$ and the true parameter ($\Psi(P_0)$) can be
represented in first order by a i.i.d. sum:
\begin{equation}\label{eqn:IC}
  \Psi(P_n) - \Psi(P_0) = \frac{1}{n} IC(O_i; \nu) + op(1/\sqrt{n})
\end{equation}
where $IC(O_i; \nu)$ is a function of the data and possibly other parameters
$\nu$. Importantly, such estimators have mean-zero Gaussian limiting
distributions; thus, in the univariate case, one has that
\begin{equation}\label{eqn:limit_dist}
  \sqrt{n}(\Psi(P_n) - \Psi(P_0)) = N(0, (IC(O_i; \nu))^2),
\end{equation}
so that inference for the estimator of interest may be obtained in terms of
the influence function. For this simple case, a 95\% confidence interval may be
derived as:
\begin{equation}\label{eqn:CI}
  \Psi(P^{\star}_n) \pm 1.96 \sqrt{\frac{\hat{\sigma}^2}{n}},
\end{equation}
where $SE=\sqrt{\frac{\hat{\sigma}^2}{n}}$ and $\hat{\sigma}^2$ is the sample
variance of the estimated IC's: $IC(O; \hat{\nu})$. One can use the functional
delta method to derive the influence curve if a parameter of interest may be
written as a function of other asymptotically linear estimators.

## The WASH Benefits Example Dataset

The data come from a study of the effect of water quality, sanitation, hand
washing, and nutritional interventions on child development in rural Bangladesh
(WASH Benefits Bangladesh): a cluster-randomised controlled trial
[@luby2018effects]. The study enrolled pregnant women in their first or second
trimester from the rural villages of Gazipur, Kishoreganj, Mymensingh, and
Tangail districts of central Bangladesh, with an average of eight women per
cluster. Groups of eight geographically adjacent clusters were block-randomised,
using a random number generator, into six intervention groups (all of which
received weekly visits from a community health promoter for the first 6 months
and every 2 weeks for the next 18 months) and a double-sized control group (no
intervention or health promoter visit). The six intervention groups were:

1. chlorinated drinking water;
2. improved sanitation;
3. handwashing with soap;
4. combined water, sanitation, and hand washing;
5. improved nutrition through counseling and provision of lipid-based nutrient
   supplements; and
6. combined water, sanitation, handwashing, and nutrition.

In the workshop, we concentrate on child growth (size for age) as the outcome of
interest. For referene, this trial was registered with ClinicalTrials.gov as
NCT01590095.

```{r load_washb_data_intro, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)

# read in data
dat <- read_csv(here("data", "washb_data.csv"))
dat
```

For the purposes of this workshop, we we start by treating the data as
independent and identically distributed (i.i.d.) random draws from a very large
target population. We could, with available options, account for the clustering
of the data (within sampled geographic units), but, for simplification, we avoid
these details in these workshop presentations, although modifications of our
methodology for biased samples, repeated measures, etc., are available.

We have 28 variables measured, of which 1 variable is set to be the outcome of
interest. This outcome, $Y$, is the weight-for-height Z-score (`whz` in `dat`);
the treatment of interest, $A$, is the randomized treatment group (`tr` in
`dat`); and the adjustment set, $W$, consists simply of *everything else*. This
results in our observed data structure being $n$ i.i.d. copies of $O_i = (W_i,
A_i, Y_i)$, for $i = 1, \ldots, n$.

### The variables

Using the [`skimr` package](https://CRAN.R-project.org/package=skimr), we can
quickly summarize the variables measured in the WASH Benefits data set:

```{r skim_washb_data, message=FALSE, warning=FALSE}
library(skimr)
skim(dat)
```

A convenient summary of the relevant variables is given just above, complete
with a small visualization describing the marginal characteristics of each
covariate. Note that the *asset* variables reflect socio-economic status of the
study participants. Notice also the uniform distribution of the treatment groups
(with twice as many controls); this is, of course, by design.
